---
title: "enviPathR"
author:
  - name: Giulio Benedetti
    affiliation: University of Turku
    email: giulio.benedetti@utu.fi
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
        self_contained: true
        code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('enviPathR')`"
vignette: >
  %\VignetteIndexEntry{enviPathR: access enviPath and predict pathways from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

username <- Sys.getenv("EP_USERNAME")
password <- Sys.getenv("EP_PASSWORD")
```

## Introduction

enviPath is a biotransformation database and pathway prediction system focusing
on xenobiotic compounds that exacerbate environmental pollution. Although the
enviPath resources are accessible via an intuitive website, this approach cannot
ensure analytical efficiency and reproducibility. For this, enviPath provides
a RESTful API through which both data retrieval and pathway prediction are
possible. The current package, enviPathR, offers an R client to the enviPath API
and thus brings most of its capabilities to R practitioners studying
environmental pollution and organic contaminant biotransformation.

## Tutorial

This section outlines how to use enviPathR for several different purposes. If
the client interface feels somewhat familiar, that is because it is inspired by
[KEGGREST](https://bioconductor.org/packages/release/bioc/html/KEGGREST.html){target="_blank"},
the R client for the KEGG RESTful API. While enabling all resource look-up and
pathway prediction functionality, the package currently lacks the API methods
for object creation and deletion. That can be added upon request (external
contributions are also welcome).

### Installation

Bioconductor release version:

```{r install-release, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("enviPathR")
```

Beta version:

```{r install-devel, eval = FALSE}
remotes::install_github("enviPath/enviPathR")
```

Then, we can import enviPathR and another couple of optional packages used in
this tutorial.

```{r setup, message = FALSE}
library(enviPathR)
library(igraph)
library(knitr)
```

All the following methods are further described in the enviPathR
[function reference](https://envipath.github.io/enviPathR/reference/index.html){target="_blank"}.

### `epTypes`: explore enviPath database structure

enviPath divides its conceptual and biological items into different object
categories, which can be navigated with `epTypes`.

```{r types}
types <- epTypes()
```

The first element of the output contains a vector of object types that can be
listed and explored with `epList`.

```{r types-listable}
# View listable object types
print(types$listable)
```

The second element is a data frame with the available links between object
types that can be retrieved with `epLink`.

```{r types-linkable}
# View linkable object types
kable(types$linkable)
```

The graph structure of the database can be visualised by converting the links to
an igraph object.

```{r types-plot}
# Convert edges data frame to igraph object
ep_graph <- graph_from_data_frame(types$linkable)

# Visualise database graph
plot(ep_graph)
```

### `epLogin`: log into an enviPath personal account

It is necessary to hold an enviPath account to use the API and thus also
enviPathR. A free (for academic use) account can be created at the [enviPath
website](https://envipath.org/){target="_blank"}. The same login credentials can
then be passed to `epLogin`, which keeps you permanently logged in unless the
cached package cookies are cleared.

```{r login}
# Log into your enviPath account
epLogin(username, password)
```

After logging in once, all remaining methods work right out the box.

### `epList`: retrieve lists of objects from enviPath

The available elements for each object type can be explored using `epList`.

```{r list-pkg}
# List available packages
pkg_df <- epList("package")

# Select reviewed packages
to_keep <- pkg_df$reviewStatus == "reviewed"
pkg_df <- pkg_df[to_keep, ]

# Remove lengthy description
pkg_df$description <- NULL

# View list of packages
kable(pkg_df)
```

By default, objects are listed from all available packages, but objects can be
listed from a specific package using the argument `pkg`. For example, it is
possible to retrieve pathways only from the EAWAG-BBD package by providing its
unique identifier.

```{r list-path}
# Select desired package
pkg_name <- "EAWAG-BBD"
pkg_id <- pkg_df$id[pkg_df$name == pkg_name]

# List pathways from desired package
path_df <- epList("pathway", pkg = pkg_id)

# View list of pathways
path_df |>
    head() |>
    kable()
```

Similarly, the other supported enviPath object types can also be explored this
way (see `epTypes`).

### `epLink`: map between different types of objects

Links between object types can be retrieved with `epLink`, which takes the
origin and target object types, and optional initial values with the argument
`init`. The latter is recommended to prevent the search from exploding.

```{r link-one}
# Define reaction of interest
rxn_id <- "2b6bbcc5-77f4-4bed-92a9-731cdc978f6a"

# Map desired reaction to related compounds (educts and products)
rxn2cpd <- epLink("reaction", "compound", init = rxn_id)

# View reaction-to-compound mapping
kable(rxn2cpd)
```

Multiple initial values can also be used, so that the output represents a
many-to-many mapping.

```{r link-many}
# Select pathways of interest
path_ids <- path_df$id[1:2]

# Map desired pathways to related compounds (nodes)
path2cpd <- epLink("pathway", "compound", init = path_ids, pkg = pkg_id)

# View pathway-to-compound mapping
kable(path2cpd)
```

Similarly, the other supported enviPath object types can also be linked this
way (see `epTypes`).

### `epGet`: fetch raw objects from enviPath

It may be useful at times to retrieve raw objects to access their original
properties. This is done with `epGet`.

```{r get-one}
# Fetch reaction object
rxn <- epGet("reaction", rxn_id)

# View part of reaction data
rxn |>
    head() |>
    print()
```

Multiple initial values can also be input to fetch a list of several different
objects.

```{r get-many}
# Fetch multiple pathways
paths <- epGet("pathway", init = path_ids, pkg = pkg_id)

# View part of pathway data
paths |>
    lapply(head) |>
    print()
```

It is also possible to extract specific properties from certain object types
with the argument `property`. These include pathway edges and nodes and compound
structures.

```{r get-property}
# Extract node property from pathway objects
nodes <- epGet("pathway", init = path_ids, pkg = pkg_id, property = "node")

# View nodes
print(nodes)
```

### `epModel`: predict compound biotransformation pathways

enviPath provides a set of pre-trained models for chemical fate prediction and
half-life persistence of both known and unknown compounds given their SMILES
representation. The available models can be retrieved with `epList`.

```{r model-setting}
# List available model settings
set_df <- epList("setting")

# View list
kable(set_df)
```

Pathway prediction is performed with `epModel`, which uses the enviFormer model
by default. The output list contains two data frames representing the nodes
(compounds) and edges (reactions) of the predicted pathway, respectively.

```{r model-former}
# Define smiles of interest
smiles <- "ClC(Cl)=C(Cl)Cl"

# Perform pathway prediction with enviFormer
former_out <- epModel(smiles)

# View compounds in predicted pathway
kable(former_out$nodes)

# View reactions in predicted pathway
kable(former_out$edges)
```

Different model settings can be specified with the argument `setting`. For
example, the PEPPER model can be used which also returns half-lime estimates for
the predicted degradation steps.

```{r model-pepper}
# Set id for PEPPER model setting
set_name <- "Global Setting - ECC and App Domain - PEPPER"
set_id <- set_df$id[set_df$name == set_name]

# Perform pathway prediction with PEPPER
pepper_out <- epModel(smiles, set_id)

# View compounds in predicted pathway
kable(pepper_out$nodes)

# View reactions in predicted pathway
kable(pepper_out$edges)
```

Remarkably, the output of `epModel` can be directly converted to an igraph
object and visualised as a pathway graph. Here, the plot function from base R
is used for simplicity, but the ggraph package is recommended for more advanced graph visualisation.

```{r model-plot}
# Convert model output to igraph object
path_graph <- graph_from_data_frame(
    pepper_out$edges,
    vertices = pepper_out$nodes
)

# Visualise predicted pathway
plot(path_graph)
```

As mentioned above, ggraph can be used to create more informative visualisations
including organised layouts and additional aesthetic layers.

```{r model-ggraph, message = FALSE}
library(ggraph)

ggraph(path_graph, layout = "sugiyama") +
    geom_edge_link(aes(colour = probability), arrow = arrow(type = "closed")) +
    geom_node_point(size = 3) +
    geom_node_text(aes(label = name), vjust = 2) +
    scale_edge_colour_continuous(limits = c(0, 1), low = "white", high = "red") +
    theme_graph()
```

## Reproducibility

### Cite enviPathR

```{r citation}
citation("enviPathR")
```

### Session info

```{r session-info}
sessionInfo()
```
